A sheaf--function dictionary for algebraic tori over local fields

### 2013-08-17 shanghai.tex
More specific citations

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves.
Since it is a rigid symmetric monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the sheaf--function dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5}, we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show that it is an isomorphism in Theorem~\ref{thm:snake}, assuming only
that the geometric component group of $G$ is finitely generated.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}, and more recently Drinfeld has referred to such objects
as character sheaves.  Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The other main use
of the term ``character sheaf'' traces back to Lusztig \cite{lusztig:85a}.  Most of his energy is devoted to
handling non-commutative groups; though he does define character sheaves for some
non-connected groups he restricts his attention to reductive groups with cyclic component group.
For extensions of $\ZZ/n\ZZ$ by a torus, it is not difficult to relate his Frobenius-stable character
sheaves to our quasicharacter sheaves.

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid symmetric monoidal category
\cite{etingof:09a} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf-hom functor 
$\sheafHom(\ - \ ,\EE)$.
The category of characters sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}.
This rigid monoidal category structure for $\QC(G)$ puts a group
structure on the set $\QCiso{G}$ of isomorphism classes in $\QC(G)$.
We now begin developing the tools to relate $\QCiso{G}$ to $\Hom(G(\Fq), \EEx)$.

@@

  Since the action of $\Gal(\bFq/\Fq)$ is continuous, there is an $m$
  so that $\Frob{\Fq}^m$ acts trivially on $X$; write $\Fqm$ for the degree
  $m$ extension of $\Fq$ and $\Gamma_m$ for $\Gal(\Fqm/\Fq)$.  Since
  $T^m-1$ factors as a squarefree product of cyclotomic polynomials,
  the Chinese remainder theorem expresses $\ZZ[\Gamma_m]$ as a product
  of $\ZZ[\zeta_d]$ for $d$ dividing $m$.  Each $\ZZ[\zeta_d]$ is the
  maximal order in a number field and thus $\ZZ[\Gamma_m]$ is a
  product of Dedekind domains.  Therefore $X$ decomposes into a direct
  sum of cyclic $\ZZ[\Gamma_m]$-modules \cite{brandal:CommRingsFGModDecompose},
  each of the \David{Add a reference to this section}
  form $\ZZ[\Gamma_m] / I$ for some ideal $I \subset \ZZ[\Gamma_m]$.
  The ideal $I$ corresponds to a direct product of ideals $I_d$ in the
  $\ZZ[\zeta_d]$, and within each $\ZZ[\zeta_d]$ the Chinese remainder
  theorem allows a further decomposition of $\ZZ[\zeta_d] / I_d$ as a
  direct sum of $\ZZ[\zeta_d]$-modules of the form $\ZZ[\zeta_d]/P^s$.
  So if we can construct, for every cyclotomic ring $\ZZ[\zeta_d]$,
  prime ideal $P \subset \ZZ[\zeta_d]$ and positive integer $s$, a
  $\psi \colon \ZZ[\zeta_d]/P^s \rightarrow \EEx$ with
  \begin{equation} \label{eq:psi-condition}
    \tilde{f}(x) = \frac{\psi(\zeta_d \cdot x)}{\psi(x)},
  \end{equation}
  then the existence of the original $\psi$ follows because $X$ is a
  direct sum of such $\ZZ[\zeta_d]/P^s$.

@@

The fundamental properties of $\GN{T}_d$ that we need are:
 \begin{enumerate}
   \labitem{(GN.0)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically.
  \labitem{(GN.1)}{GNd} $\GN{T}_d$ is a smooth, commutative group scheme,
  locally of finite type over $\Fq$, and the component group of $\GN{T}_d$ is
  geometrically finitely generated. (In fact, $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$,
  and the latter is finitely generated by \cite{xarles:93a}.)
  \labitem{(GN.2)}{d+1} 
  \end{enumerate}

@@

The techniques in this paper apply to all non-archimdean local fields $K$ with residue field $k$,
and to all algebraic tori $T$ over $K$, without placing any restrictions on the characteristic of $K$
or on the ramification of $T$. Because of this, it is natural to ask if we can compare
quasicharacter sheaves on $T$ and on $T'$ under conditions when there is a
natural comparison between characters of $T(K)$ and of $T'(K')$.
This comparison of has been made precise in \cite{chai-yu:01a},
building on ideas of \cite{deligne:84a}.

@@

This paper has shown how quasicharacter sheaves 
on the Greenberg transform of the N\/eron model of $T$ 
determine admissible characters of $T(K)$, and vice versa, functorially.
When coupled with the reciprocity map $\text{rec}_T $ for $T$, 
which enjoys the same functorial properties, 
it follows that quasicharacter sheaves for $T$ 
determine Langlands parameters for $T$, and vice versa, functorially.
\[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr}  & \\
\Hom_{<\infty}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell) \\
\end{tikzcd}
\]
We suspect that it is possible to extract Langlands parameters 
from quasicharacter sheaves directly and that the reciprocity map 
can then be be viewed as pairing characters and parameters coming
 from the same quasicharacter sheaf. 
 Results in this paper provide tools for reduction the problem to the case $T=\Gm{K}$. 
 In that case, our suspicions are all but confirmed 
 by the class field theory of Serre-Hazewinkel
 \citelist{\cite{serre:60a},\cite{serre:61a},\cite{demazure-gabriel:GroupesAlgebriques}*{Appendice}},
 especially as revisited in \cite{suzuki-yoshida:12a}. 
 What remains to be done is a careful comparison of 
 quasicharacter sheaves for $\Gm{K}$ and certain local systems on
  the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{suzuki-yoshida:12a}.

>>>>

In Part 1 of this paper we study the category $\QC(G)$ of quasicharacter sheaves.
Since it is a rigid symmetric monoidal category
the set of isomorphism classes $\QCiso{G}$ forms a group.
Motivated by the sheaf--function dictionary for commutative, connected
algebraic groups \cite{deligne:SGA4.5},\David{more specific citation} we explore the trace of Frobenius morphism
\[
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\]
and show that it is an isomorphism in Theorem~\ref{thm:snake}, assuming only
that the geometric component group of $G$ is finitely generated.

@@

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}, \David{more specific citation} and more recently Drinfeld has referred to such objects
as character sheaves.  Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The other main use
of the term ``character sheaf'' traces back to Lusztig \cite{lusztig:85a}*{Def.~2.10}.  Most of his energy is devoted to
handling non-commutative groups; though he does define character sheaves for some
non-connected groups he restricts his attention to reductive groups with cyclic component group.
For extensions of $\ZZ/n\ZZ$ by a torus, it is not difficult to relate his Frobenius-stable character
sheaves to our quasicharacter sheaves.

@@

The category $\QC(G)$ of quasicharacter sheaves on $G$ is a rigid symmetric monoidal category
\cite{etingof:09a}*{\S~1.10} under the tensor product
$\qcs{L} \otimes \qcs{L'} \ceq (\gqcs{L}\otimes\gqcs{L'}, \mu\otimes\mu', \phi\otimes \phi')$,
with duals given by applying the sheaf-hom functor 
$\sheafHom(\ - \ ,\EE)$.
The category of characters sheaves on $G$ is not abelian, so $\QC(G)$
is not a tensor category in the sense of \cite{deligne:02a}. \David{more specific citation}
This rigid monoidal category structure for $\QC(G)$ puts a group
structure on the set $\QCiso{G}$ of isomorphism classes in $\QC(G)$.
We now begin developing the tools to relate $\QCiso{G}$ to $\Hom(G(\Fq), \EEx)$.

@@

  Since the action of $\Gal(\bFq/\Fq)$ is continuous, there is an $m$
  so that $\Frob{\Fq}^m$ acts trivially on $X$; write $\Fqm$ for the degree
  $m$ extension of $\Fq$ and $\Gamma_m$ for $\Gal(\Fqm/\Fq)$.  Since
  $T^m-1$ factors as a squarefree product of cyclotomic polynomials,
  the Chinese remainder theorem expresses $\ZZ[\Gamma_m]$ as a product
  of $\ZZ[\zeta_d]$ for $d$ dividing $m$.  Each $\ZZ[\zeta_d]$ is the
  maximal order in a number field and thus $\ZZ[\Gamma_m]$ is a
  product of Dedekind domains.  Therefore $X$ decomposes into a direct
  sum of cyclic $\ZZ[\Gamma_m]$-modules \cite{brandal:CommRingsFGModDecompose}, \David{more specific citation}
  each of the \David{Add a reference to this section}
  form $\ZZ[\Gamma_m] / I$ for some ideal $I \subset \ZZ[\Gamma_m]$.
  The ideal $I$ corresponds to a direct product of ideals $I_d$ in the
  $\ZZ[\zeta_d]$, and within each $\ZZ[\zeta_d]$ the Chinese remainder
  theorem allows a further decomposition of $\ZZ[\zeta_d] / I_d$ as a
  direct sum of $\ZZ[\zeta_d]$-modules of the form $\ZZ[\zeta_d]/P^s$.
  So if we can construct, for every cyclotomic ring $\ZZ[\zeta_d]$,
  prime ideal $P \subset \ZZ[\zeta_d]$ and positive integer $s$, a
  $\psi \colon \ZZ[\zeta_d]/P^s \rightarrow \EEx$ with
  \begin{equation} \label{eq:psi-condition}
    \tilde{f}(x) = \frac{\psi(\zeta_d \cdot x)}{\psi(x)},
  \end{equation}
  then the existence of the original $\psi$ follows because $X$ is a
  direct sum of such $\ZZ[\zeta_d]/P^s$.

@@

The fundamental properties of $\GN{T}_d$ that we need are:
 \begin{enumerate}
   \labitem{(GN.0)}{dFq} $\GN{T}_d(\Fq) = \TT(\OK/\pK^d)$, canonically.
  \labitem{(GN.1)}{GNd} $\GN{T}_d$ is a smooth, commutative group scheme,
  locally of finite type over $\Fq$, and the component group of $\GN{T}_d$ is
  geometrically finitely generated. (In fact, $\pi_0(\GN{T}_d) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$,
  and the latter is finitely generated by \cite{xarles:93a}*{2.18}.)
  \labitem{(GN.2)}{d+1} 
  \end{enumerate}

@@

The techniques in this paper apply to all non-archimdean local fields $K$ with residue field $k$,
and to all algebraic tori $T$ over $K$, without placing any restrictions on the characteristic of $K$
or on the ramification of $T$. Because of this, it is natural to ask if we can compare
quasicharacter sheaves on $T$ and on $T'$ under conditions when there is a
natural comparison between characters of $T(K)$ and of $T'(K')$.
This comparison of has been made precise in \cite{chai-yu:01a}, \David{more specific citation?}
building on ideas of \cite{deligne:84a}.

@@

This paper has shown how quasicharacter sheaves 
on the Greenberg transform of the N\/eron model of $T$ 
determine admissible characters of $T(K)$, and vice versa, functorially.
When coupled with the reciprocity map $\text{rec}_T $ for $T$, 
which enjoys the same functorial properties, 
it follows that quasicharacter sheaves for $T$ 
determine Langlands parameters for $T$, and vice versa, functorially.
\[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr}  & \\
\Hom_{<\infty}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell) \\
\end{tikzcd}
\]
We suspect that it is possible to extract Langlands parameters 
from quasicharacter sheaves directly and that the reciprocity map 
can then be be viewed as pairing characters and parameters coming
 from the same quasicharacter sheaf. 
 Results in this paper provide tools for reduction the problem to the case $T=\Gm{K}$. 
 In that case, our suspicions are all but confirmed 
 by the class field theory of Serre-Hazewinkel
 \citelist{\cite{serre:60a},\cite{serre:61a},\cite{demazure-gabriel:GroupesAlgebriques}*{Appendice}},
 especially as revisited in \cite{suzuki-yoshida:12a}*{\S~2}. 
 What remains to be done is a careful comparison of 
 quasicharacter sheaves for $\Gm{K}$ and certain local systems on
  the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{suzuki-yoshida:12a}.
